---
title: "Cyclistic Bike Ride Project"
author: "Cyclistic Bike Ride"
date: "2023-05-12"
output:
  html_document: default
  pdf_document: default
---

PROJECT GOAL:- To determine bike rides stats based on **Seasons** for the year *2021.*

## Installing required packages

```{r Install Packages, message=FALSE, warning=FALSE}
install.packages("tidyverse",repos = "http://cran.us.r-project.org")
install.packages("lubridate",repos = "http://cran.us.r-project.org")
install.packages("ggplot2",repos = "http://cran.us.r-project.org")
install.packages("ggthemes",repos = "http://cran.us.r-project.org")
```

## Loading the packages

```{r Load the packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(ggthemes)
```

ASK PHASE: To determine the goal of the project, by asking the project manager
we got to know that we are required to extract the seasonality of bike ride data
based on ride numbers and average duration of ride.

PREPARE: 

## Import and combine required datasets
We are provided with the 12 .csv files each containing data of one
month. We combine these datasets to form a full year dataset.

```{r Import and combine datasets, message=FALSE, warning=FALSE}
df1 <- read.csv('202101-divvy-tripdata.csv')
df2 <- read.csv('202102-divvy-tripdata.csv')
df3 <- read.csv('202103-divvy-tripdata.csv')
df4 <- read.csv('202104-divvy-tripdata.csv')
df5 <- read.csv('202105-divvy-tripdata.csv')
df6 <- read.csv('202106-divvy-tripdata.csv')
df7 <- read.csv('202107-divvy-tripdata.csv')
df8 <- read.csv('202108-divvy-tripdata.csv')
df9 <- read.csv('202109-divvy-tripdata.csv')
df10 <- read.csv('202110-divvy-tripdata.csv')
df11 <- read.csv('202111-divvy-tripdata.csv')
df12 <- read.csv('202112-divvy-tripdata.csv')

bike_ride <- rbind(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)
```

PROCESS:

## Cleaning the dataset
We observe many missing values in datasets, and we assign them to NA's.

```{r Assign NA Values}
bike_ride[bike_ride == "" | bike_ride == " "] <- NA
```

After contacting the project manager for the concern of NA values in dataset, she
instructed us to remove the rows with NA values to make the data consistent.

```{r Remove NA values, message=FALSE, warning=FALSE}
bike_ride <- drop_na(bike_ride)
```

Initially we had 5595063 rows and after omitting rows with NA values we have 
4588302 rows.

We then format the dataset with appropriate formatting.
*ride_length* and *start_month* column is created.

```{r Format Dataset, message=FALSE, warning=FALSE}
bike_ride$rideable_type <- as.factor(bike_ride$rideable_type)
bike_ride$member_casual <- as.factor(bike_ride$member_casual)
bike_ride$started_at <- as_datetime(bike_ride$started_at)
bike_ride$ended_at <- as_datetime(bike_ride$ended_at)
bike_ride$start_month <- months.Date(bike_ride$started_at)
bike_ride$ride_length <- difftime(bike_ride$ended_at,bike_ride$started_at,units = c("mins"))
```

ANALYZE:

## Analyzing the dataset

Selecting the required columns, based on that average duration and number of rides
is calculated.

```{r Analyze dataset, warning=FALSE}
bike_ride <- select(bike_ride, ride_id, rideable_type, started_at, ended_at, member_casual, start_month, ride_length)
bike_ride$ride_length <- as.numeric(bike_ride$ride_length)
bike_ride <- subset(bike_ride, ride_length > 0 & ride_length < 160)
bike_ride$ride_length <- round(bike_ride$ride_length,2)
bike_ride$started_at <- as.character(bike_ride$started_at)
bike_ride$start_date <- as.Date(bike_ride$started_at)
bike_ride <- arrange(bike_ride, start_date, decreasing = FALSE)
bike_ride_1 <- bike_ride %>% 
  group_by(member_casual,start_date) %>% 
  summarise(number_of_rides=n(),Average_ride_Length=mean(ride_length))
```

SHARE:

## Visualizing data 

We will make use of ggplot and ggthemes for data visualization.
First we will Visualize data on Time Period vs Number of Rides.

```{r Visualizing data, warning=FALSE}
ggplot(data=bike_ride_1) + 
  geom_point(mapping = aes(x = start_date, y = number_of_rides),alpha = 0.3) + 
  facet_wrap(~member_casual) + 
  geom_smooth(mapping = aes(x = start_date, y = number_of_rides)) + theme_economist() +
  ggtitle("Monthly data of bike rides for the year 2021",subtitle = "Data viz for casual & Member(Paid) Users") +
  labs(x = "Time Period(Months)",y = "Number of rides")
```

Secondly we will visualize data on Time Period vs Average ride length

```{r Visualize data, warning=FALSE}
ggplot(data=bike_ride_1) + 
  geom_line(mapping = aes(x = start_date, y = Average_ride_Length,color = member_casual),size = 1) + 
  theme_economist() + 
  ggtitle("Average duration of bike rides for the year 2021",subtitle = "Data Viz for casual & Member(Paid) Users") +
  labs(x = "Time Period(Months)",y = "Average Ride Duration (Mins)")
```

ACT:

We can clearly see that the Casual Users of bike ride are more scattered around the 
month of *April* to *November*. Also the ride duration done by Casual users is
considerably more than the Membership users. Therefor company should restrict the 
duration time for casual riders, after which they should be ask for plan upgrade
for further riding.

